---
title: "AMS 580 Final Exam Review Session"
author: "Weihao Wang"
date: "Spring 2026"
output:
  pdf_document:
    toc: true
  word_document:
    toc: true
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---
Part 1
Classification task
logistic/ SVM/ RF/ DT/ NN/ Boosting
confusion matrix/ accuracy

Part 2
Regression task
LR/ RF/ DT/ NN/ Boosting
MSE/ RMSE


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Review Session Goals

This file is designed as a compact final-exam review for AMS 580. It reviews the most important modeling workflows, common R functions, and common mistakes.

Main topics:

- Linear regression and model diagnostics
- Logistic regression and classification metrics
- Stepwise and best-subset variable selection
- Penalized regression: ridge, lasso, elastic net
- PCA, correspondence analysis, and clustering
- CART, random forest, boosting, SVM
- Neural networks for regression and classification
- RNN/LSTM concepts and reproducibility issues

Most examples use the `banknote` and `Titanic2` datasets.

# 0. Package Loading and Function Masking

```{r packages}
# IMPORTANT: Load packages carefully. Some packages have functions with the same names.
# Common masking examples:
# 1. dplyr::select() can be masked by MASS::select().
# 2. dplyr::filter() can be confused with stats::filter().
# 3. dplyr::lag() can be confused with stats::lag().
# 4. keras3 and tensorflow depend on a correctly initialized Python environment.
# 5. randomForest::importance() and xgboost importance functions are different.

# Good practice: use package::function() when there is possible conflict.
# Example: dplyr::select(data, x, y), MASS::stepAIC(model), e1071::svm(...)

required_packages <- c(
  "tidyverse", "caret", "MASS", "leaps", "glmnet", "rpart", "rpart.plot",
  "randomForest", "e1071", "pROC", "nnet", "xgboost", "factoextra"
)

# Do not install packages during an exam unless explicitly allowed.
# If a package is missing, install it before the review/exam.
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message("Package not installed: ", pkg)
  }
}

library(tidyverse)
library(caret)
```

```{r masking-demo, eval=FALSE}
# If you load MASS after tidyverse, MASS::select may mask dplyr::select.
library(MASS)

# This may fail if select() is masked:
# select(mydata, x1, x2)

# Safer version:
#dplyr::select(mydata, x1, x2)

# Use conflicts() to check masked functions.
conflicts()
```

# 1. Load and Prepare Data

```{r load-data}
# Put this Rmd file in the same folder as banknote(2).csv and Titanic2.csv.
# Common mistake: file path errors during knitting.
# Always set the working directory correctly or use relative paths.

banknote <- read.csv("banknote.csv")
titanic <- read.csv("Titanic2.csv")

str(banknote)
str(titanic)
```

```{r data-cleaning}
# Banknote: class is coded as 0/1. For classification models, convert it to a factor.
# Common mistake: leaving class as numeric makes some models treat it as regression.

banknote <- banknote %>%
  mutate(
    class = factor(class, levels = c(0, 1), labels = c("genuine", "forged"))
  )

# Titanic: Survived is also 0/1. Convert to factor for classification.
# Remove variables that are mostly identifiers or high-cardinality text fields.
# Common mistake: including Name or Ticket directly can create too many meaningless dummy variables.

Titanic_clean <- titanic %>%
  dplyr::select(Survived, Pclass, Sex, Age, SibSp, Parch, Fare, Embarked) %>%
  mutate(
    Survived = factor(Survived, levels = c(0, 1), labels = c("No", "Yes")),
    Pclass = factor(Pclass),
    Sex = factor(Sex),
    Embarked = factor(Embarked)
  ) %>%
  drop_na()

summary(banknote)
summary(Titanic_clean)
```

```{r train-test-split}
set.seed(123)

# Common mistake: splitting without set.seed() makes results change every time.
# Common mistake: preprocessing the full dataset before splitting may cause data leakage.
# For simple examples, we split first and then estimate preprocessing parameters from training data.

idx_bank <- caret::createDataPartition(banknote$class, p = 0.8, list = FALSE)
bank_train <- banknote[idx_bank, ]
bank_test  <- banknote[-idx_bank, ]

idx_titanic <- caret::createDataPartition(Titanic_clean$Survived, p = 0.8, list = FALSE)
titanic_train <- Titanic_clean[idx_titanic, ]
titanic_test  <- Titanic_clean[-idx_titanic, ]

prop.table(table(bank_train$class))
prop.table(table(bank_test$class))
prop.table(table(titanic_train$Survived))
prop.table(table(titanic_test$Survived))
```

# 2. Evaluation Metrics

## 2.1 Regression metrics

```{r regression-metrics}
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))
mae <- function(y, yhat) mean(abs(y - yhat))
r_squared <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

# Common mistake: using training RMSE only. Always report test/validation RMSE when evaluating prediction.
```

## 2.2 Classification metrics

```{r classification-metrics}
# Confusion matrix convention matters.
# caret::confusionMatrix(data = predicted, reference = truth, positive = "Yes")
# Common mistake: reversing predicted and reference changes interpretation.

classification_report <- function(pred, truth, positive_level) {
  caret::confusionMatrix(data = pred, reference = truth, positive = positive_level)
}
```

# 3. Linear Regression

Use Titanic data to predict `Fare` as a numeric response. This is only for reviewing the linear regression workflow.

```{r linear-regression}
# Common mistake: lm() is for continuous response, not binary classification.
# If the response is 0/1, use glm(..., family = binomial), not lm().

lm_train <- titanic_train %>%
  mutate(logFare = log(Fare + 1))

lm_test <- titanic_test %>%
  mutate(logFare = log(Fare + 1))

lm_fit <- lm(logFare ~ Pclass + Sex + Age + SibSp + Parch + Embarked, data = lm_train)
summary(lm_fit)

lm_pred <- predict(lm_fit, newdata = lm_test)
rmse(lm_test$logFare, lm_pred)
mae(lm_test$logFare, lm_pred)
r_squared(lm_test$logFare, lm_pred)
```

```{r linear-diagnostics}
# Diagnostic plots check linearity, equal variance, outliers, and normality of residuals.
# Common mistake: a significant coefficient does not automatically mean strong prediction.
# Common mistake: R-squared is not a test-set prediction metric unless computed on test data.

par(mfrow = c(2, 2))
plot(lm_fit)
par(mfrow = c(1, 1))
```

Key points:

- Use `lm()` for continuous response.
- Interpret coefficients conditional on the other variables.
- Check residuals and outliers.
- Use test RMSE/MAE for prediction performance.
- Do not compare models only by training error.

# 4. Logistic Regression

Use Titanic survival as the binary response.

```{r logistic-regression}
# Logistic regression models log-odds.
# Common mistake: using predict(model) directly gives log-odds, not probabilities.
# Use type = "response" to get probabilities.

logit_fit <- glm(Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked,
                 data = titanic_train,
                 family = binomial)
summary(logit_fit)

prob_yes <- predict(logit_fit, newdata = titanic_test, type = "response")

# Common mistake: threshold = 0.5 is default but not always optimal, especially for imbalanced data.
pred_yes <- ifelse(prob_yes > 0.5, "Yes", "No")
pred_yes <- factor(pred_yes, levels = levels(titanic_test$Survived))

classification_report(pred_yes, titanic_test$Survived, positive_level = "Yes")
```

```{r logistic-roc}
# ROC/AUC uses probabilities, not class labels.
# Common mistake: passing predicted classes to ROC instead of probabilities.

if (requireNamespace("pROC", quietly = TRUE)) {
  roc_obj <- pROC::roc(response = titanic_test$Survived,
                       predictor = prob_yes,
                       levels = c("No", "Yes"))
  pROC::auc(roc_obj)
  plot(roc_obj, main = "ROC Curve: Logistic Regression")
}
```

Important interpretation:

- Coefficients are on the log-odds scale.
- `exp(coef)` gives odds ratios.
- Accuracy alone can be misleading for imbalanced data.
- Sensitivity and specificity depend on the threshold.

```{r odds-ratio}
exp(coef(logit_fit))
```

# 5. Stepwise and Best-Subset Variable Selection

```{r stepwise-selection}
# Stepwise selection is based on AIC by default when using MASS::stepAIC().
# Common mistake: loading MASS can mask dplyr::select(). Use MASS::stepAIC explicitly.
# Common mistake: selecting variables on the full dataset before train/test split causes data leakage.

if (requireNamespace("MASS", quietly = TRUE)) {
  full_lm <- lm(logFare ~ Pclass + Sex + Age + SibSp + Parch + Embarked, data = lm_train)
  null_lm <- lm(logFare ~ 1, data = lm_train)

  step_fit <- MASS::stepAIC(null_lm,
                            scope = list(lower = null_lm, upper = full_lm),
                            direction = "both",
                            trace = FALSE)
  summary(step_fit)

  step_pred <- predict(step_fit, newdata = lm_test)
  rmse(lm_test$logFare, step_pred)
}
```

```{r best-subset-selection}
# Best subset selection tries combinations of predictors.
# Common mistake: best subset can be computationally expensive for many predictors.
# Common mistake: the selected model may overfit if selection is not validated.

if (requireNamespace("leaps", quietly = TRUE)) {
  best_fit <- leaps::regsubsets(logFare ~ Pclass + Sex + Age + SibSp + Parch + Embarked,
                                data = lm_train,
                                nvmax = 6)
  best_sum <- summary(best_fit)
  best_sum$adjr2
  best_sum$bic

  plot(best_sum$bic, type = "b",
       xlab = "Number of variables",
       ylab = "BIC",
       main = "Best Subset Selection by BIC")
}
```

Exam reminders:

- AIC tends to choose larger models than BIC.
- Stepwise selection is greedy; it does not guarantee the global best model.
- Best subset is more exhaustive but can be unstable.
- Always evaluate selected models on validation/test data.

# 6. Penalized Regression: Ridge, Lasso, Elastic Net

Use `glmnet`. It requires a numeric matrix `x` and response vector `y`.

```{r glmnet-regression}
# Common mistake: glmnet does not accept a regular data frame with factors directly.
# Use model.matrix() to create dummy variables.
# Common mistake: model.matrix() includes an intercept column by default. Remove it with [, -1].
# Common mistake: standardization is important for penalized regression. glmnet standardizes by default.

if (requireNamespace("glmnet", quietly = TRUE)) {
  x_train <- model.matrix(logFare ~ Pclass + Sex + Age + SibSp + Parch + Embarked,
                          data = lm_train)[, -1]
  y_train <- lm_train$logFare

  x_test <- model.matrix(logFare ~ Pclass + Sex + Age + SibSp + Parch + Embarked,
                         data = lm_test)[, -1]
  y_test <- lm_test$logFare

  set.seed(123)
  ridge_cv <- glmnet::cv.glmnet(x_train, y_train, alpha = 0) # alpha = 0 ridge
  lasso_cv <- glmnet::cv.glmnet(x_train, y_train, alpha = 1) # alpha = 1 lasso
  enet_cv  <- glmnet::cv.glmnet(x_train, y_train, alpha = 0.5) # elastic net

  ridge_pred <- predict(ridge_cv, newx = x_test, s = "lambda.min")
  lasso_pred <- predict(lasso_cv, newx = x_test, s = "lambda.min")
  enet_pred  <- predict(enet_cv,  newx = x_test, s = "lambda.min")

  c(
    ridge_rmse = rmse(y_test, ridge_pred),
    lasso_rmse = rmse(y_test, lasso_pred),
    enet_rmse  = rmse(y_test, enet_pred)
  )

  coef(lasso_cv, s = "lambda.min")
}
```

```{r glmnet-classification}
# For binary classification in glmnet, use family = "binomial".
# Response can be a factor or 0/1 numeric, but be careful about positive class interpretation.

if (requireNamespace("glmnet", quietly = TRUE)) {
  x_train_b <- model.matrix(class ~ ., data = bank_train)[, -1]
  y_train_b <- bank_train$class
  x_test_b  <- model.matrix(class ~ ., data = bank_test)[, -1]

  set.seed(123)
  lasso_bin <- glmnet::cv.glmnet(x_train_b, y_train_b,
                                 family = "binomial",
                                 alpha = 1)

  prob_forged <- predict(lasso_bin, newx = x_test_b,
                         s = "lambda.min", type = "response")
  pred_forged <- ifelse(prob_forged > 0.5, "forged", "genuine")
  pred_forged <- factor(pred_forged, levels = levels(bank_test$class))

  classification_report(pred_forged, bank_test$class, positive_level = "forged")
}
```

Conceptual reminders:

- Ridge shrinks coefficients but usually does not set them exactly to zero.
- Lasso can perform variable selection by setting coefficients to zero.
- Elastic net combines ridge and lasso.
- `lambda.min` gives the minimum cross-validation error.
- `lambda.1se` gives a simpler model within one standard error of the minimum.

# 7. PCA and Clustering

Use only numeric predictors from the banknote dataset.

```{r pca}
# PCA is unsupervised. Do not use the class label when computing PCs.
# Common mistake: including the response variable in PCA causes leakage.
# Common mistake: forgetting scale. PCA is sensitive to variable scale.

bank_x <- banknote %>%
  dplyr::select(variance, skewness, curtosis, entropy)

pca_fit <- prcomp(bank_x, center = TRUE, scale. = TRUE)
summary(pca_fit)

plot(pca_fit, type = "l", main = "Scree Plot")
biplot(pca_fit, scale = 0)
```

```{r clustering}
# K-means is also unsupervised.
# Common mistake: comparing cluster labels directly to class labels without allowing label switching.
# Cluster 1 does not necessarily mean genuine or forged.

set.seed(123)
bank_scaled <- scale(bank_x)

km2 <- kmeans(bank_scaled, centers = 2, nstart = 25)
table(Cluster = km2$cluster, TrueClass = banknote$class)

# Elbow method
wss <- sapply(1:8, function(k) {
  kmeans(bank_scaled, centers = k, nstart = 25)$tot.withinss
})
plot(1:8, wss, type = "b",
     xlab = "Number of clusters K",
     ylab = "Total within-cluster sum of squares",
     main = "Elbow Method")
```

Important reminders:

- PCA is for dimension reduction and visualization.
- K-means assumes roughly spherical clusters.
- Hierarchical clustering does not require choosing K first, but you still need to cut the tree.
- Scaling is usually necessary for PCA and distance-based clustering.

# 8. CART: Classification and Regression Trees

```{r cart-classification}
# CART is easy to interpret but can overfit.
# Common mistake: not pruning the tree.
# Common mistake: using type = "vector" or default predictions when class labels/probabilities are needed.

if (requireNamespace("rpart", quietly = TRUE)) {
  cart_fit <- rpart::rpart(class ~ ., data = bank_train, method = "class")
  printcp(cart_fit)

  if (requireNamespace("rpart.plot", quietly = TRUE)) {
    rpart.plot::rpart.plot(cart_fit)
  }

  cart_pred <- predict(cart_fit, newdata = bank_test, type = "class")
  classification_report(cart_pred, bank_test$class, positive_level = "forged")
}
```

```{r cart-pruning}
if (requireNamespace("rpart", quietly = TRUE)) {
  # Choose cp with lowest cross-validation error.
  best_cp <- cart_fit$cptable[which.min(cart_fit$cptable[, "xerror"]), "CP"]
  pruned_cart <- rpart::prune(cart_fit, cp = best_cp)

  if (requireNamespace("rpart.plot", quietly = TRUE)) {
    rpart.plot::rpart.plot(pruned_cart)
  }

  pruned_pred <- predict(pruned_cart, newdata = bank_test, type = "class")
  classification_report(pruned_pred, bank_test$class, positive_level = "forged")
}
```

Conceptual reminders:

- A single tree has high variance.
- Pruning reduces overfitting.
- Trees handle nonlinear interactions naturally.
- Trees do not require feature scaling.

# 9. Random Forest

```{r random-forest}
# Random forest reduces variance by averaging many trees.
# Common mistake: thinking random forest is automatically interpretable like one tree.
# Common mistake: using importance as causal explanation. Variable importance is predictive, not causal.

if (requireNamespace("randomForest", quietly = TRUE)) {
  set.seed(123)
  rf_fit <- randomForest::randomForest(class ~ .,
                                       data = bank_train,
                                       ntree = 500,
                                       mtry = 2,
                                       importance = TRUE)
  rf_fit

  rf_pred <- predict(rf_fit, newdata = bank_test, type = "class")
  classification_report(rf_pred, bank_test$class, positive_level = "forged")

  randomForest::varImpPlot(rf_fit, main = "Random Forest Variable Importance")
}
```

Key points:

- Bagging uses bootstrap samples.
- Random forest adds random feature selection at each split.
- Important tuning parameters: `ntree`, `mtry`, node size / tree depth.
- RF often performs well but is less interpretable than a single tree.

# 10. Boosting: AdaBoost, Gradient Boosting, XGBoost

```{r xgboost-banknote}
# XGBoost requires numeric matrix input.
# Common mistake: passing a factor response directly to xgboost.
# Common mistake: labels for binary logistic objective should be 0/1 numeric.
# Common mistake: using best_nrounds when the object stores best_iteration or best_iter.

if (requireNamespace("xgboost", quietly = TRUE)) {
  x_train_xgb <- as.matrix(bank_train %>% dplyr::select(-class))
  y_train_xgb <- ifelse(bank_train$class == "forged", 1, 0)
  x_test_xgb <- as.matrix(bank_test %>% dplyr::select(-class))
  y_test_xgb <- ifelse(bank_test$class == "forged", 1, 0)

  dtrain <- xgboost::xgb.DMatrix(data = x_train_xgb, label = y_train_xgb)
  dtest  <- xgboost::xgb.DMatrix(data = x_test_xgb, label = y_test_xgb)

  params <- list(
    objective = "binary:logistic",
    eval_metric = "logloss",
    max_depth = 3,
    eta = 0.1,
    subsample = 0.8,
    colsample_bytree = 0.8
  )

  set.seed(123)
  xgb_fit <- xgboost::xgb.train(
    params = params,
    data = dtrain,
    nrounds = 100,
    watchlist = list(train = dtrain, test = dtest),
    verbose = 0
  )

  xgb_prob <- predict(xgb_fit, newdata = dtest)
  xgb_pred <- ifelse(xgb_prob > 0.5, "forged", "genuine")
  xgb_pred <- factor(xgb_pred, levels = levels(bank_test$class))

  classification_report(xgb_pred, bank_test$class, positive_level = "forged")
}
```

Conceptual reminders:

- AdaBoost sequentially updates observation weights.
- Gradient boosting sequentially fits the negative gradient/residual direction.
- XGBoost improves gradient boosting with regularization, shrinkage, subsampling, and efficient implementation.
- Boosting is order-dependent; later trees correct previous errors.
- Tune `nrounds`, `eta`, `max_depth`, `subsample`, and `colsample_bytree`.

# 11. Support Vector Machine

```{r svm-banknote}
# SVM is sensitive to feature scale. Scale numeric predictors.
# Common mistake: using a linear kernel when the boundary is nonlinear, or using RBF without tuning cost/gamma.
# Common mistake: probability = TRUE must be set during training if you want probabilities later.

if (requireNamespace("e1071", quietly = TRUE)) {
  set.seed(123)
  svm_fit <- e1071::svm(class ~ ., data = bank_train,
                        kernel = "radial",
                        cost = 1,
                        gamma = 0.5,
                        scale = TRUE,
                        probability = TRUE)

  svm_pred <- predict(svm_fit, newdata = bank_test)
  classification_report(svm_pred, bank_test$class, positive_level = "forged")
}
```

```{r svm-tuning, eval=FALSE}
# Tuning example. This may take time.
# Common mistake: tuning on the test set. Use cross-validation inside the training set.

set.seed(123)
svm_tune <- e1071::tune.svm(
  class ~ ., data = bank_train,
  kernel = "radial",
  cost = c(0.1, 1, 10),
  gamma = c(0.01, 0.1, 1)
)
summary(svm_tune)
```

Conceptual reminders:

- Linear SVM seeks a maximum-margin hyperplane.
- Kernel SVM maps data into a higher-dimensional feature space implicitly.
- Larger `cost` penalizes misclassification more strongly and can overfit.
- RBF `gamma` controls how local the decision boundary is.

# 12. Neural Networks for Classification

Use `nnet` for a lightweight exam-review example.

```{r nnet-classification}
# Neural networks usually require scaled numeric features.
# Common mistake: not scaling inputs, causing unstable training.
# Common mistake: expecting identical results every run without setting seed.
# Common mistake: too many hidden units with small data can overfit.

if (requireNamespace("nnet", quietly = TRUE)) {
  bank_train_scaled <- bank_train
  bank_test_scaled <- bank_test

  pp_bank <- caret::preProcess(bank_train_scaled %>% dplyr::select(-class),
                               method = c("center", "scale"))
  bank_train_scaled[, 1:4] <- predict(pp_bank, bank_train_scaled %>% dplyr::select(-class))
  bank_test_scaled[, 1:4]  <- predict(pp_bank, bank_test_scaled %>% dplyr::select(-class))

  set.seed(123)
  nn_fit <- nnet::nnet(class ~ ., data = bank_train_scaled,
                       size = 5,
                       decay = 0.01,
                       maxit = 500,
                       trace = FALSE)

  nn_pred <- predict(nn_fit, newdata = bank_test_scaled, type = "class")
  nn_pred <- factor(nn_pred, levels = levels(bank_test_scaled$class))

  classification_report(nn_pred, bank_test_scaled$class, positive_level = "forged")
}
```

Conceptual reminders:

- Hidden layers learn nonlinear transformations of the input features.
- Activation functions introduce nonlinearity.
- Regularization can be done by weight decay, dropout, early stopping, or smaller networks.
- Neural networks may be less interpretable than regression/tree models.

# 13. Neural Networks for Regression

```{r nnet-regression}
# Example: predict logFare in Titanic.
# Common mistake: neural networks for regression need linear output, but nnet handles this with linout = TRUE.
# Common mistake: response scaling may help, especially for larger neural networks.

if (requireNamespace("nnet", quietly = TRUE)) {
  # model.matrix handles categorical predictors.
  x_nn_train <- model.matrix(logFare ~ Pclass + Sex + Age + SibSp + Parch + Embarked,
                             data = lm_train)[, -1]
  x_nn_test <- model.matrix(logFare ~ Pclass + Sex + Age + SibSp + Parch + Embarked,
                            data = lm_test)[, -1]

  pp_nn <- caret::preProcess(as.data.frame(x_nn_train), method = c("center", "scale"))
  x_nn_train_s <- predict(pp_nn, as.data.frame(x_nn_train))
  x_nn_test_s  <- predict(pp_nn, as.data.frame(x_nn_test))

  nn_reg_train <- cbind(logFare = lm_train$logFare, x_nn_train_s)
  nn_reg_test  <- cbind(logFare = lm_test$logFare, x_nn_test_s)

  set.seed(123)
  nn_reg_fit <- nnet::nnet(logFare ~ ., data = nn_reg_train,
                           size = 4,
                           decay = 0.01,
                           linout = TRUE,
                           maxit = 500,
                           trace = FALSE)

  nn_reg_pred <- predict(nn_reg_fit, newdata = nn_reg_test)
  rmse(nn_reg_test$logFare, nn_reg_pred)
}
```

# 14. RNN and LSTM Review

RNN/LSTM models are mainly for sequential data. Banknote and Titanic are tabular data, so they are not natural examples for RNN/LSTM. Use time series or ordered sequences instead.

```{r rnn-lstm-concept, eval=FALSE}
# Conceptual input shape for keras3 RNN/LSTM:
# x should be a 3D array: samples × timesteps × features
# Example: dim(x_train) = c(number_of_sequences, sequence_length, number_of_features)

# Common mistake: passing a 2D matrix to LSTM.
# Common mistake: wrong shape order. Keras uses samples, timesteps, features.
# Common mistake: using future information when creating sequences for time-series prediction.
# Common mistake: random train/test split for time series. Usually use chronological split.

library(keras3)

seq_len <- 10
n_features <- 1

model_lstm <- keras_model_sequential() |>
  layer_input(shape = c(seq_len, n_features)) |>
  layer_lstm(units = 32) |>
  layer_dense(units = 1)

model_lstm |>
  compile(
    optimizer = "adam",
    loss = "mse"
  )
```

## RNN vs LSTM conceptual differences

- Simple RNN has one hidden-state update mechanism.
- LSTM has gates: input gate, forget gate, output gate, and cell state.
- LSTM is designed to handle longer-term dependencies better than simple RNN.
- Both require sequence-shaped input.
- For small/simple sequences, RNN may be enough; for longer dependency, LSTM often works better.

## Reproducibility warning for keras3

```{r keras-reproducibility, eval=FALSE}
# keras3::set_random_seed(123) is necessary but may not be sufficient.
# GPU operations can still be nondeterministic.
# For stronger reproducibility, set determinism and consider CPU-only execution.

Sys.setenv(CUDA_VISIBLE_DEVICES = "-1")  # run before TensorFlow is loaded, optional CPU-only mode

library(keras3)
library(tensorflow)

keras3::set_random_seed(123)
tf$config$experimental$enable_op_determinism()

# Common mistake: setting seed after model construction.
# Set seed before creating the model and before fitting.
# Common mistake: using validation_split with shuffle = TRUE when exact reproducibility is needed.
```

# 15. Model Comparison Summary

```{r model-comparison-template, eval=FALSE}
# A clean model comparison table should use test-set metrics.
# Common mistake: comparing training accuracy of one model with test accuracy of another model.
# Common mistake: changing train/test split across models.

results <- tibble::tibble(
  Model = c("Logistic Regression", "CART", "Random Forest", "SVM", "XGBoost", "Neural Network"),
  Accuracy = c(NA, NA, NA, NA, NA, NA),
  Sensitivity = c(NA, NA, NA, NA, NA, NA),
  Specificity = c(NA, NA, NA, NA, NA, NA),
  AUC = c(NA, NA, NA, NA, NA, NA)
)

results
```

General comparison ideas:

- Linear/logistic regression: interpretable baseline.
- Penalized regression: useful for many predictors and multicollinearity.
- CART: interpretable but unstable.
- Random forest: strong nonlinear baseline, less interpretable.
- Boosting/XGBoost: often highly accurate, requires tuning.
- SVM: strong for medium-size data, sensitive to scaling and tuning.
- Neural network: flexible but may require more tuning and data.
- RNN/LSTM: for sequential/time-series data, not ordinary tabular data.

# 16. Final Exam Common Mistake Checklist

## Data preparation

- Do not include the response variable in scaling/PCA/clustering predictors.
- Split the data before estimating preprocessing parameters.
- Convert binary classification responses to factors when using classification models.
- Do not include ID variables or high-cardinality text fields without proper processing.
- Check missing values before modeling.

## Modeling

- Use `lm()` for continuous response.
- Use `glm(..., family = binomial)` for binary logistic regression.
- Use `type = "response"` for logistic regression probabilities.
- Use `type = "class"` for CART class predictions.
- Use `model.matrix()` for `glmnet`, `xgboost`, and some neural network workflows.
- Scale inputs for PCA, clustering, SVM, neural networks, and penalized regression.
- Do not scale tree-based models because it is usually unnecessary.

## Evaluation

- Use test-set or cross-validation performance.
- Accuracy is not enough if the classes are imbalanced.
- Sensitivity and specificity require defining the positive class.
- ROC/AUC should be computed from probabilities or scores, not hard class labels.
- For regression, report RMSE/MAE/test R-squared.

## Function names and masking

- Use `dplyr::select()` if `MASS` is loaded.
- Use `MASS::stepAIC()` for stepwise AIC selection.
- Use `e1071::svm()` for SVM.
- Use `randomForest::randomForest()` and `randomForest::varImpPlot()` for random forest.
- Use `glmnet::cv.glmnet()` for penalized regression.
- Use `rpart::rpart()` and `rpart.plot::rpart.plot()` for CART.

## Interpretation

- Prediction performance is not the same as causal explanation.
- Variable importance is not causality.
- A statistically significant variable may not improve prediction much.
- A more complicated model is not always better.
- Always connect the method choice to the data type and prediction goal.

# 17. Very Short Conceptual Review

- **Bias-variance tradeoff:** simple models may underfit; complex models may overfit.
- **Cross-validation:** estimates out-of-sample performance and helps tune parameters.
- **Regularization:** controls model complexity.
- **PCA:** finds directions of maximum variance, unsupervised.
- **Clustering:** groups observations without labels.
- **Bagging:** averages many models to reduce variance.
- **Boosting:** sequentially improves weak learners to reduce errors.
- **SVM:** maximizes margin; kernels allow nonlinear boundaries.
- **Neural networks:** learn nonlinear feature transformations.
- **LSTM:** sequence model designed to remember longer-term dependencies.



