---
title: "Boosting Methods with the Banknote Data"
author: "Standard Solutions"
output:
  pdf_document:
    toc: true
    toc_depth: '2'
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, message=FALSE, warning=FALSE}
library(caret)
library(adabag) # adaboost
library(gbm) # gradient boost
library(xgboost) # xgboost
library(dplyr)
library(ggplot2)

set.seed(123)
```

## Data preparation

```{r}
# Read the data
banknote <- read.csv("banknote.csv")

# Make sure the response is a factor for classification.
# We keep the positive class as "1" (forged) throughout the exercise.
banknote$class <- factor(banknote$class, levels = c(0, 1))

# Split into training and testing sets using the same split for all models.
set.seed(123)
idx <- createDataPartition(banknote$class, p = 0.75, list = FALSE)
train.data <- banknote[idx, ]
test.data  <- banknote[-idx, ]

nrow(train.data)
nrow(test.data)
```

The outputs above report the numbers of observations in the training and testing sets.

## Q2. AdaBoost

```{r}
# AdaBoost from the adabag package.
# mfinal = 100 means we use 100 weak learners.
# coeflearn = "Breiman" uses the classical Breiman-style updating rule.
set.seed(123)
ada_model <- boosting(
  class ~ ., data = train.data,
  mfinal = 100,   # 100 trees
  coeflearn = "Breiman",  # update rule
  boos = TRUE   # bootstrap samples
)

ada_pred <- predict(ada_model, newdata = test.data)
ada_class <- factor(ada_pred$class, levels = c(0, 1))

cm_ada <- confusionMatrix(ada_class, test.data$class, positive = "1")
cm_ada
```

```{r}
ada_sens <- cm_ada$byClass["Sensitivity"]
ada_spec <- cm_ada$byClass["Specificity"]
ada_acc  <- cm_ada$overall["Accuracy"]

ada_sens
ada_spec
ada_acc
```

## Q3. Gradient Boosting (GBM)

```{r}
# gbm expects a numeric 0/1 response for Bernoulli loss.
train_gbm <- train.data %>% mutate(class_num = as.numeric(as.character(class)))
test_gbm  <- test.data  %>% mutate(class_num = as.numeric(as.character(class)))

# Grid of tuning values requested in the exercise.
gbm_grid <- expand.grid(
  interaction.depth = c(1, 2, 3),  # the depth of each tree
  shrinkage = c(0.01, 0.05, 0.1),  # learning rate
  n.minobsinnode = c(5, 10)  # the minimum number of observations in terminal nodes
)

# We use a reasonably large maximum number of trees, then let CV choose the best one.
# The best model is the one with the smallest CV error at its own best tree count.
gbm_results <- data.frame()
gbm_models <- list()

for (i in 1:nrow(gbm_grid)) {
  set.seed(123)
  fit_i <- gbm(
    formula = class_num ~ variance + skewness + curtosis + entropy,
    data = train_gbm,
    distribution = "bernoulli",
    n.trees = 2000, # maximum number of boosting trees
    interaction.depth = gbm_grid$interaction.depth[i],
    shrinkage = gbm_grid$shrinkage[i],
    n.minobsinnode = gbm_grid$n.minobsinnode[i],
    cv.folds = 5,  # 5-fold cross-validation
    bag.fraction = 0.8, # each tree is trained using a random 80% subsample of the training data
    verbose = FALSE
  )
  best_iter <- gbm.perf(fit_i, method = "cv", plot.it = FALSE)
  # This finds the best number of trees based on cross-validation error
  best_cv <- fit_i$cv.error[best_iter]
  gbm_results <- rbind(
    gbm_results,
    data.frame(
      interaction.depth = gbm_grid$interaction.depth[i],
      shrinkage = gbm_grid$shrinkage[i],
      n.minobsinnode = gbm_grid$n.minobsinnode[i],
      best_n.trees = best_iter,
      cv_error = best_cv
    )
  )
  gbm_models[[i]] <- fit_i
}

gbm_results <- gbm_results[order(gbm_results$cv_error), ]
gbm_results
```

```{r}
# Extract the best GBM fit and its best number of trees.
best_gbm_row <- gbm_results[1, ]
best_gbm_index <- which(
  gbm_grid$interaction.depth == best_gbm_row$interaction.depth &
  gbm_grid$shrinkage == best_gbm_row$shrinkage &
  gbm_grid$n.minobsinnode == best_gbm_row$n.minobsinnode
)[1]

best_gbm_model <- gbm_models[[best_gbm_index]]
best_gbm_trees <- best_gbm_row$best_n.trees
best_gbm_row
```

```{r}
# Predict probabilities, then convert to class labels using 0.5.
gbm_prob <- predict(best_gbm_model, newdata = test_gbm, n.trees = best_gbm_trees, type = "response")
gbm_class <- factor(ifelse(gbm_prob > 0.5, 1, 0), levels = c(0, 1))

cm_gbm <- confusionMatrix(gbm_class, test.data$class, positive = "1")
cm_gbm
```

```{r}
gbm_sens <- cm_gbm$byClass["Sensitivity"]
gbm_spec <- cm_gbm$byClass["Specificity"]
gbm_acc  <- cm_gbm$overall["Accuracy"]

gbm_sens
gbm_spec
gbm_acc
```

## Q4. XGBoost

```{r}
# xgboost requires a numeric design matrix and numeric 0/1 labels.
train_x <- as.matrix(train.data[, c("variance", "skewness", "curtosis", "entropy")])
test_x  <- as.matrix(test.data[,  c("variance", "skewness", "curtosis", "entropy")])
train_y <- as.numeric(as.character(train.data$class))
test_y  <- as.numeric(as.character(test.data$class))
# XGBoost training feature matrix

dtrain <- xgb.DMatrix(data = train_x, label = train_y)
dtest  <- xgb.DMatrix(data = test_x,  label = test_y)

xgb_grid <- expand.grid(
  eta = c(0.05, 0.1),
  max_depth = c(2, 3, 4),
  subsample = c(0.8, 1.0),
  colsample_bytree = c(0.8, 1.0)
)

# eta is the learning rate.
# max_depth controls the maximum depth of each tree.
# subsample controls the fraction of training observations used for each boosting round.
# colsample_bytree controls the fraction of predictors used for each tree.

xgb_results <- data.frame()

for (i in 1:nrow(xgb_grid)) {
  params_i <- list(
    objective = "binary:logistic",
    eval_metric = "logloss",
    eta = xgb_grid$eta[i],
    max_depth = xgb_grid$max_depth[i],
    subsample = xgb_grid$subsample[i],
    colsample_bytree = xgb_grid$colsample_bytree[i]
  )

#  objective = "binary:logistic" means binary classification with logistic output probabilities.
#  eval_metric = "logloss" means the model is evaluated using log loss.

  set.seed(123)
  cv_i <- xgb.cv(
    params = params_i,
    data = dtrain,
    nrounds = 300,
    nfold = 5,
    early_stopping_rounds = 20,
    verbose = 0
  )

#  nrounds = 300 means the maximum number of boosting rounds is 300.
#  nfold = 5 means 5-fold CV.
#  early_stopping_rounds = 20 means training stops if the validation log loss does not improve for 20 rounds.
#  verbose = 0 suppresses printed training output.
  
  xgb_results <- rbind(
    xgb_results,
    data.frame(
      eta = xgb_grid$eta[i],
      max_depth = xgb_grid$max_depth[i],
      subsample = xgb_grid$subsample[i],
      colsample_bytree = xgb_grid$colsample_bytree[i],
      best_nrounds = cv_i$best_iteration,
      best_logloss = cv_i$evaluation_log$test_logloss_mean[cv_i$best_iteration]
    )
  )
}

xgb_results <- xgb_results[order(xgb_results$best_logloss), ]
xgb_results
```

```{r}
best_xgb_row <- xgb_results[1, ]
best_xgb_row
```

```{r}
# Train the final XGBoost model using the best settings selected by CV.
best_params <- list(
  objective = "binary:logistic",
  eval_metric = "logloss",
  eta = best_xgb_row$eta,
  max_depth = best_xgb_row$max_depth,
  subsample = best_xgb_row$subsample,
  colsample_bytree = best_xgb_row$colsample_bytree
)

set.seed(123)
final_xgb <- xgb.train(
  params = best_params,
  data = dtrain,
  nrounds = best_xgb_row$best_nrounds,
  verbose = 0
)

xgb_prob <- predict(final_xgb, newdata = dtest)
xgb_class <- factor(ifelse(xgb_prob > 0.5, 1, 0), levels = c(0, 1))

cm_xgb <- confusionMatrix(xgb_class, test.data$class, positive = "1")
cm_xgb
```

```{r}
xgb_sens <- cm_xgb$byClass["Sensitivity"]
xgb_spec <- cm_xgb$byClass["Specificity"]
xgb_acc  <- cm_xgb$overall["Accuracy"]

xgb_sens
xgb_spec
xgb_acc
```

## Q5. Variable importance

```{r}
# GBM variable importance
# summary(..., plotit = FALSE) returns a table of relative influence.
gbm_importance <- summary(best_gbm_model, n.trees = best_gbm_trees, plotit = FALSE)
gbm_importance
```

```{r}
# XGBoost variable importance
xgb_importance <- xgb.importance(
  feature_names = colnames(train_x),
  model = final_xgb
)
xgb_importance
```

```{r}
# Plot GBM importance
ggplot(gbm_importance, aes(x = reorder(var, rel.inf), y = rel.inf)) +
  geom_col() +
  coord_flip() +
  labs(title = "GBM Variable Importance", x = "Variable", y = "Relative Influence") +
  theme_minimal()
```

```{r}
# Plot XGBoost importance using Gain
xgb_importance_plot <- xgb_importance[, c("Feature", "Gain")]

ggplot(xgb_importance_plot, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col() +
  coord_flip() +
  labs(title = "XGBoost Variable Importance (Gain)", x = "Variable", y = "Gain") +
  theme_minimal()
```

The most influential variables are the ones near the top of the importance rankings above. Because importance is model-specific, the exact ranking may differ across boosting methods.

## Q6. Final comparison of AdaBoost, Gradient Boosting, and XGBoost

```{r}
comparison_table <- data.frame(
  Model = c("AdaBoost", "Gradient Boosting", "XGBoost"),
  Sensitivity = c(as.numeric(ada_sens), as.numeric(gbm_sens), as.numeric(xgb_sens)),
  Specificity = c(as.numeric(ada_spec), as.numeric(gbm_spec), as.numeric(xgb_spec)),
  Accuracy = c(as.numeric(ada_acc), as.numeric(gbm_acc), as.numeric(xgb_acc))
)

comparison_table
```

```{r}
best_model <- comparison_table[which.max(comparison_table$Accuracy), ]
best_model
```

All the same!

In general, XGBoost often performs strongly because it combines gradient boosting with regularization, shrinkage, subsampling, and efficient tree construction. However, the strongest method in a given application still depends on the tuning choices, sample size, noise level, and structure of the predictors.
