---
title: "Review of R: Lecture 4"
author: "Weihao Wang"
output: pdf_document
---

**1. Regression and correlation**

1.1 Simple linear regression

The linear regression model is given by
y = alpha + beta*x + e
in which the e are assumed independent and normal(0, sig^2), The parameters alpha, beta and sig. can be estimated using the method of least squares. In particular, the values of alpha and beta can be obtained by minimizing the sum of squared residuals, and sig. can be estimated via the sum of squared residuals.

```{r}
library(ISwR)
data(thuesen) 
attach(thuesen) # blood -- x velocity -- y
lm(short.velocity~blood.glucose) # fit linear regression: lm(y~x)
# short.velocity = 1.09781 + 0.02196 * blood.glucose
summary(lm(short.velocity~blood.glucose))
plot(blood.glucose,short.velocity) 
abline(lm(short.velocity~blood.glucose))
```

1.2 Residuals and fitted values

We have seen how summary can be used to extract information about the results of a regression analysis. Two further extraction functions are fitted and resid.
```{r}
lm.velo<-lm(short.velocity~blood.glucose) 
short.velocity
# fitted yi = a + b*xi
fitted(lm.velo)
resid(lm.velo) 
plot(blood.glucose,short.velocity) 
abline(lm.velo)

#plot(blood.glucose,short.velocity) 
#lines(blood.glucose,fitted(lm.velo)) # Becasue of NAs

plot(blood.glucose,short.velocity)
lines(blood.glucose[!is.na(short.velocity)],fitted(lm.velo))
# blood.glucose[!is.na(short.velocity) & !is.na(blood.glucose)] #
# lines(blood.glucose[!is.na(short.velocity) &is.na(blood.glucose)],fitted(lm.velo))

options(na.action=na.exclude) 
lm.velo<-lm(short.velocity~blood.glucose) 
fitted(lm.velo)
plot(blood.glucose,short.velocity)
abline(lm.velo) 
segments(blood.glucose,fitted(lm.velo),blood.glucose,short.velocity)
plot(fitted(lm.velo),resid(lm.velo)) 
qqnorm(resid(lm.velo))
```

Exercise
```{r}
logret<- read.table("Lecture1_6logreturn.txt", header=T)
names(logret)
attach(logret)

(fit1 <- lm(Pfizer~Intel))
summary(fit1)
names(fit1)
fit1$coefficients
plot(Intel, Pfizer)
abline(lm(Pfizer~Intel))

fit2 <- lm(Pfizer~-1+Intel) # regression without intercept
summary(fit2) # y = b*x

fitted(fit1)
resid(fit1)
plot(Intel, Pfizer)
lines(Intel, fitted(fit1))
```

To fit a simple linear regression of Pfizer ~ Intel step by step:

```{r}
# Step 1: Fit the Simple Linear Regression Model
model <- lm(Pfizer ~ Intel, data = logret)
summary(model) # Summary of the model

# Step 2: Check Assumptions
# 2.1. Linearity
plot(logret$Intel, logret$Pfizer, main="Scatter plot of Pfizer vs Intel", xlab="Intel", ylab="Pfizer")
abline(model, col="red")
# 2.2. Constant Variability (Homoscedasticity)
plot(model$fitted.values, model$residuals, 
     main="Residuals vs Fitted",
     xlab="Fitted values", ylab="Residuals")
abline(h = 0, col = "red") # Residuals vs Fitted values plot
# 2.3. Nearly Normal Residuals
qqnorm(model$residuals) # Q-Q plot for normality
qqline(model$residuals, col="red")
hist(model$residuals, breaks=20, main="Histogram of Residuals", xlab="Residuals")


plot(model) # diagnosis

# If the assumptions are violated (e.g., non-linearity or non-constant variability), you might need to transform one or both variables. Common transformations include log, square root, or inverse.
# Log transformation on both variables
model_sq <- lm((Pfizer)^2 ~ (Intel)^2, data = logret)
# Check the summary of the transformed model
summary(model_sq)
```

1.3 Correlation

Correlation is a statistical measure that describes the strength and direction of the relationship between two variables. It quantifies how much two variables move together or in opposite directions.

Range: The correlation coefficient typically ranges from -1 to +1.  
+1: A perfect positive correlation, meaning that as one variable increases, the other also increases proportionally.  
-1: A perfect negative correlation, meaning that as one variable increases, the other decreases proportionally.  
0: No correlation, indicating no linear relationship between the variables. 

The function 'cor' can be used to compute the correlation between two or more vectors.
```{r}
attach(thuesen)
cor(blood.glucose,short.velocity) 
cor(blood.glucose,short.velocity,use="complete.obs") 
# use: an optional character string giving a method for computing covariances in the presence of missing values. 
cor(thuesen,use="complete.obs") 
cor.test(blood.glucose,short.velocity)

# 1. Pearson Correlation Coefficient: Measures the linear relationship between two continuous variables. It assumes that the variables are normally distributed.
# 2. Spearman Rank Correlation: A non-parametric measure that assesses how well the relationship between two variables can be described using a monotonic function (useful when the data does not meet the assumptions for Pearson correlation).

# Compute Pearson correlation (default)
cor(blood.glucose,short.velocity,use="complete.obs", method = "pearson")
# Compute Spearman correlation
cor(blood.glucose,short.velocity,use="complete.obs", method = "spearman")

# Scatter plot to visualize correlation
plot(blood.glucose,short.velocity, main="Scatter plot of Pfizer vs Intel", xlab="Intel", ylab="Pfizer")
```

Exercise
```{r}
cor(Intel, Pfizer)
cor.test(Intel, Pfizer)
```

**2. Muliple linear regression & General linear model**

Multiple linear regression is used to model the relationship between one numeric outcome or response or dependent variable (Y), and several (multiple) explanatory or independent or predictor or regressor variables (X). When some predictors are categorical variables, we call the subsequent regression model as the General Linear Model.

2.1 Import Data

```{r}
# Directly from the internet
# data <-read.table('http://www.randomservices.org/random/data/Galton.txt', header = T)
# From your own directories
# You need to set your directory at the beginning
data <- read.table("Lecture4_Galton.txt", header = TRUE)
# Or use the whole directory
# data <- read.table("/Users/ian/Desktop/SBU/AMS 394/summer 2021/Galton.txt", header = TRUE)

#Recall:
#Y = height of child 
#x1 = height of father
#x2 = height of mother 
#x3 = gender of children
y<-data$Height 
x1<-data$Father
x2<-data$Mother 
x3<-rep(1, dim(data)[1])
x3[data$Gender == "M"]<-0
```

2.2 Multiple regression using the lm() function

```{r}
#To perform Multiple Regression,
#we use the same functions as we use in Simple Linear Regression 
#Notice that we use "+" between two variables in lm()
mod<-lm(y ~ x1+x2+x3)
summary(mod)
#Notice that 4 p-values are very small,
#which means variables x1,x2,x3
#have strong linear relationship with y.
#We conclude that all beta are significantly different from zero.
#Since F = 529>2.615, we reject Ho,
#and conclude that our model predicts height better than by chance.
#Equivalently, F-statistic's p-value: < 2.2e-16, hence we reject Ho.
```

2.3 Obtain confidence intervals for model parameters

```{r}
#The following function is used to get CI of your "Beta"
confint(mod,level=0.95) 
#confint(mod,conf.level=0.95)
```

2.4 Check model goodness-of-fit

```{r}
par(mfrow=c(2,2)) 
plot(mod)
# 1 -- independence
# 2 -- normality
```

2.5 General Linear Model using the function

We can use glm() as well – this is especially convenient when we have **categorical variables** in our data set. Instead of creating dummy variables by ourselves, R can directly work with the categorical var iables. This is in the same spirit as the Proc GLM procedure in SAS.

glm {stats}
Fitting Generalized Linear Models
Description:
glm is used to fit generalized linear models, specified by giving a symbolic description of the linear predictor and a description of the error distribution.

Usage:
glm(formula, family = gaussian, data, weights, subset,
 na.action, start = NULL, etastart, mustart, offset,
 control = list(...), model = TRUE, method = "glm.fit",
 x = FALSE, y = TRUE, contrasts = NULL, ...)
 You can see the details by help(glm)
```{r}
x3<- data$Gender
mod1<-glm(y~x1+x2+factor(x3))
summary(mod1)
#The result is similar to summary(mod)
#As we see, R uses “F” as x3's reference level because “F” comes bef ore “M” in the alphabetic order.

#Now we change it into “M”
x3<-relevel(factor(x3),ref="M")
mod1<-glm(y~x1+x2+factor(x3)) 
summary(mod1)
```

2.6 Import data in other formats

Instead of read.table(), we can use read.csv() to read .csv files, or read.xlsx to read .xlsx files. (we need to install and library the package ‘xlsx’ to read excel files)
```{r}
#install.packages("xlsx")
library(xlsx)
data1<-read.xlsx("Lecture4_heat.xlsx", 1)
data1
```

2.7 Best subset variable selection

The **regsubsets** function from the leaps package in R is used for model selection by identifying the **best subset of predictors** for linear regression models. This function **examines all possible models for a given set of predictors** and helps you choose the best combination based on criteria like the adjusted R², BIC, Cp, and others.

```{r}
# Now we need to first install the library ‘leaps’, and then we call it:
# install.packages("leaps")
library(leaps)
attach(data1)
#The attach command above will enable R to use the variables in the dataset directly.
leaps1<-regsubsets(Y~X1+X2+X3+X4,data=data1)
summary(leaps1)
leaps1<-regsubsets(Y~X1+X2+X3+X4,data=data1, nbest = 10) # nbest = 10: Selects the best 10 models for each number of predictors.
summary(leaps1)

plot(leaps1, scale="r2")
plot(leaps1, scale="adjr2") 
plot(leaps1, scale="bic") 
plot(leaps1, scale="Cp")
# Mallows' Cp is a statistical measure used in linear regression to help in model selection. It provides a trade-off between the complexity of a regression model (i.e., the number of predictors) and the quality of its fit to the data.
```

Example:

```{r}
# Example dataset
data(mtcars)
# Fit all subsets regression using regsubsets function
fit <- regsubsets(mpg ~ ., data = mtcars, nbest = 1, nvmax = NULL, method = "exhaustive")
# method = "exhaustive": Searches all possible models. Other options include "backward", "forward", and "seqrep" for different search methods.
# Summary of the regsubsets object
summary(fit)
```

2.8 Stepwise variable selection

The **step** function in the MASS package (also available in base R) is used for **stepwise model selection** in regression analysis. It **adds or removes predictors** based on criteria like AIC (Akaike Information Criterion) to find the best-fitting model.

AIC (Akaike Information Criterion): AIC balances model fit and complexity. A lower AIC is better.

Collinearity: Ensure predictors are not highly correlated before using stepwise selection, as it can lead to instability in the model.

```{r}
#Next, We use step() to perform Stepwise Regression
#step selects the model by AIC
#step is a slightly simplified version of stepAIC in package MASS
library(MASS)
step(lm(Y~X1+X2+X3+X4), data=data1)
summary(step(lm(Y~X1+X2+X3+X4), data=data1))
# The final model is: Y = beta0 +beta1X1 +beta3X3 + e
```

Forward Selection: Starts with no predictors and adds them one by one based on improvement in the model's performance.  
Backward Elimination: Starts with all predictors and removes them one by one if they don’t significantly contribute to the model.  
Both: A combination of forward and backward selection.  

Example:

```{r}
# Fit a full model with all predictors
full_model <- lm(mpg ~ ., data = mtcars)
summary(full_model)

# Perform stepwise selection
step_model <- step(full_model, direction = "both")
summary(step_model)
# 1.Initial Model: The step function starts with the full model.
# 2.Model Selection: It iteratively removes or adds predictors based on the AIC criterion. AIC balances the model fit (likelihood) with model complexity (number of predictors).
# 3.Final Model: After running through all combinations, step provides the model with the best AIC.
```

2.9 Different variable selction criteria


Please also note that SAS and R may give you different results in variable selection because different selection criteria maybe used. For example, in SAS, for stepwise variable selection, we use the F-test/Partial correlation. However, in R, we use the AIC criterion. For those of you who are highly interested in this topic, you can study on your own, for example, the following papers:

https://stat.ethz.ch/R-manual/R-devel/library/stats/html/step.html
http://analytics.ncsu.edu/sesug/2007/SA05.pdf

The AIC (Akaike information criterion) was proposed by Dr. Hirotugu Akaike.
Let L be the maximum value of the likelihood function for the given model, and let k be the number of estimated parameters in the model. The AIC value of the model is: AIC = 2k - 2ln(L)
Given a set of candidate models for the given data, the preferred mo del is the one with the minimum AIC value. AIC rewards goodness of f it (as assessed by the likelihood function), while including a penal ty against overfitting (big k).

AIC (Akaike Information Criterion) and BIC (Bayesian Information Criterion) are both metrics used for model selection in statistical analysis. They help to choose the best model from a set of models by balancing goodness-of-fit with model complexity.

**AIC** is typically preferred when the focus is on **prediction** or when the **sample size is relatively small**. It is more flexible and can handle situations where the "true" model may not be in the set of candidate models.  
**BIC** is often used when the goal is to identify the **true model** or when the **sample size is large**. BIC is more conservative, often favoring simpler models that are less likely to overfit the data.

AIC focuses on the trade-off between model fit and complexity, penalizing the number of parameters lightly.  
BIC also trades off fit and complexity, but with a stronger penalty for model complexity, especially as sample size increases.


2.10

```{r}
#install.packages("MASS")   # For stepwise regression
#install.packages("caTools") # For splitting data

# Load necessary libraries
library(MASS)
library(caTools)

# Load dataset
data <- read.csv("Lecture4_example.csv")

# Set seed for reproducibility
set.seed(42)
# Split the data
split <- sample.split(data$Y, SplitRatio = 0.8)
train_data <- subset(data, split == TRUE)
test_data <- subset(data, split == FALSE)

# Fit a full model using all features
full_model <- lm(Y ~ X1 + X2 + X3 + X4 + X5, data = train_data)

# Perform stepwise regression
stepwise_model <- stepAIC(full_model, direction = "both")
stepwise_model2 <- step(full_model, direction = "both", k = 2)

# Summary of the stepwise model
summary(stepwise_model)
summary(stepwise_model2) # the same, but you can set k=log(n) for using BIC

plot(step_model)

# Predicting the target variable (Y) on the test data
predictions <- predict(stepwise_model, newdata = test_data)

# Prediction MSE (Mean Squared Error) and Prediction R square (Coefficient of Determination) are metrics used to evaluate the performance of a predictive model, particularly in regression analysis. They provide insight into how well a model predicts unseen data. 

# Prediction MSE is the average of the squared differences between the actual and predicted values on a test dataset (data not used in model training). It measures the overall prediction error of the model.

# Prediction R^2 indicates the proportion of variance in the actual data that is predictable from the model. It measures how well the model's predictions align with the true outcomes. R^2 ranges from 0 to 1, where 1 indicates perfect prediction and 0 means the model predicts no better than the mean.

# Calculate Mean Squared Error (MSE)
mse <- mean((test_data$Y - predictions)^2)

# Calculate R-squared
ss_total <- sum((test_data$Y - mean(test_data$Y))^2)
ss_residual <- sum((test_data$Y - predictions)^2)
r_squared <- 1 - (ss_residual / ss_total)

# Print MSE and R-squared
print(paste("Mean Squared Error:", round(mse, 2)))
print(paste("R-squared:", round(r_squared, 2)))


library(ggplot2)

# Create a data frame with actual and predicted values for visualization
results <- data.frame(Actual = test_data$Y, Predicted = predictions)

# Plot actual vs predicted values using ggplot2
ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_point(color = 'blue') +
  geom_abline(slope = 1, intercept = 0, color = 'red', linetype = 'dashed') +
  labs(title = "Actual vs Predicted Values",
       x = "Actual Values",
       y = "Predicted Values") +
  theme_minimal()
```

MSE focuses on the average error in the model's predictions, with larger errors being penalized more.

R-squared tells you how much of the variability in the data is captured by the model's predictors.

MSE is a common metric used to evaluate the accuracy of a regression model. It measures the average of the squared differences between the actual values (from the test dataset) and the predicted values (from the model).
Lower MSE means the model's predictions are closer to the actual values, indicating better performance.
Higher MSE suggests larger differences between predicted and actual values, indicating poorer performance.

R-squared is a statistical measure that indicates the proportion of the variance in the dependent variable (Y) that is explained by the independent variables in the model.
R^2 = 1: The model perfectly explains all the variance in the dependent variable.
R^2 = 0: The model explains none of the variance in the dependent variable.
Higher R² indicates that the model explains more of the variance, meaning the predictions are closer to the actual values.

