---
title: "Linear Regression"
author: "Your name"
output:
  pdf_document: default
  word_document: default
---

# Install packages 

```{r}
library(MASS)
library(tidyverse)
library(caret)
library(glmnet)
```

## Exercise 1

```{r}
data <- read.csv('Galton.csv')
data$Midparent = (data$Father + data$Mother) / 2
head(data)
# characteristic variables should be transformed into dummy variables if using the lm() function
```

### (a) Predict daughter’s height using the midparent height


```{r}
daughters = subset(data, Gender == "F")
head(daughters)

model1 <- lm(Height ~ Midparent, data=daughters)
summary(model1)

daughters$pred <- model1 %>% predict(daughters)
head(daughters)

# compare Height (observed y) and pred (predicted y)
sse_daughters <- sum((daughters$Height - daughters$pred)^2)
print(sse_daughters)

mse_daughters <-sse_daughters/(nrow(daughters)-2)
print(mse_daughters)
print(nrow(daughters)-2)
# Mean Squared Error (MSE) in regression is used to measure the average squared difference between the actual and predicted values.
# It quantifies the model’s performance, with lower values indicating better predictive accuracy.
rmse_dughters <- sqrt(mse_daughters)
print(rmse_dughters)
# A useful benchmark is to compare MSE with the variance of the target variable or to compute Root Mean Squared Error (RMSE) to interpret errors in the original unit.
```

### (b) Predict son’s height using the midparent height.

```{r}
sons = subset(data, Gender == "M")
head(sons)

model2 <- lm(Height ~ Midparent, data=sons)
summary(model2)

sons$pred <- model2 %>% predict(sons)
head(sons)

sse_sons <- sum((sons$Height - sons$pred)^2)
print(sse_sons)

mse_sons <-sse_sons/(nrow(sons)-2)
print(mse_sons)
print(nrow(sons)-2)

rmse_sons <- sqrt(mse_sons)
print(rmse_sons)
```

### Exercise 2 general linear model 

```{r}
new_data <- data[, c("Father", "Mother", "Gender", "Height")]

#rename the M and F to 1 and 0
new_data$Gender=ifelse(new_data$Gender=="M",1,0)
head(new_data)
new_data$Gender<-factor(new_data$Gender)
head(new_data)

modelGLM <-glm(Height ~., data=new_data)
summary(modelGLM)

new_data$pred <- modelGLM %>% predict(new_data)

new_daughters= subset(new_data, Gender == 0)
new_sons= subset(new_data, Gender == 1)

sse_glm_daughters <- sum((new_daughters$Height - new_daughters$pred)^2)
print(sse_glm_daughters)
mse_glm_daughters <-sse_glm_daughters/(nrow(new_daughters)-2)
print(mse_glm_daughters)
sqrt(mse_glm_daughters)

sse_glm_sons <- sum((new_sons$Height - new_sons$pred)^2)
print(sse_glm_sons)
mse_glm_sons <-sse_glm_sons/(nrow(new_sons)-2)
print(mse_glm_sons)
sqrt(mse_glm_sons)
```

### Exercise 3: Comparison of the two models for estimating daughters' height
#### We can compare the SSE (or MSE or RMSE) between the two models
#### GLM is better than the simple linear regression model because 1705.501 < 1715.684

### Exercise 4: Comparison of the two models for estimating sons' height
#### We can compare the SSE (or MSE or RMSE) between the two models
#### GLM is better than the simple linear regression model because 2443.661 < 2449.981
