---
title: "Model Diagnosis"
author: "Weihao Wang"
date: "2025-02-10"
output:
  pdf_document: default
  html_document: default
---

## Model Diagnostics for Linear Regression in R

To check the assumptions and performance of a linear regression model, we typically assess:

1. Linearity
2. Homoscedasticity (Constant Variance)
3. Normality of Residuals
4. Independence of Errors
5. Multicollinearity

```{r}
library(MASS)
library(tidyverse)
library(caret)
library(glmnet)
library(leaps)

# Use variables except G1 and G2 to predicto G3
# y: G3
# x: all the other variables
data <- read.csv('math.csv', sep = ';')
data <- subset(data, select = - c(G1,G2))
str(data)

set.seed(123)
training.samples <- data$G3 %>% createDataPartition(p = 0.75, list = FALSE) # caret pkg
# Uses createDataPartition() from the caret package to split the data into training (75%) and test (25%) sets.
# The %>% operator (pipe) is from the tidyverse pkg
train.data  <- data[training.samples, ]
test.data <- data[-training.samples, ]
```

```{r}
fit <- lm(G3 ~ ., data = train.data)
```

# A. Residual Plots for Linearity & Homoscedasticity

1. Residuals should be randomly scattered (no clear pattern).  
2. The spread should be consistent across all fitted values.
```{r}
plot(fit, which = 1)  # Residuals vs Fitted
# If categorical variables are treated as numeric, the model may create artificial "steps" in residuals, leading to parallel lines.
# If some predictors take only a few distinct values, residuals will align along specific horizontal levels.
# If some predictors are highly correlated, it can cause systematic patterns in residuals.
```

# B. Check Normality of Residuals

Residuals should follow a normal distribution for valid hypothesis tests.

```{r}
plot(fit, which = 2)  # Normal Q-Q plot
# both ends of the plot fall below the straight diagonal line, this indicates heavy-tailed (platykurtic) residuals
# Residuals may have Low Kurtosis ---> Try a Box-Cox transformation to adjust distribution
# or use rlm() robust regression from MASS
fit_robust <- rlm(G3 ~ ., data = train.data)
summary(fit_robust) # robust regression methods reduce the influence of extreme values

shapiro.test(residuals(fit))
# Residuals deviate from normality (consider transformations).
```

# C. Check Homoscedasticity (Constant Variance)

Variance of residuals should be constant across fitted values.

```{r}
library(lmtest)
bptest(fit)  # Breusch-Pagan test against heteroskedasticity
# p > 0.05 → No heteroscedasticity (good).
# p < 0.05 → Heteroscedasticity detected (consider transformations like log or Box-Cox).
# E.g.
fit_trans <- lm(log(G3) ~ ., data = train.data)
# If the dependent variable has a skewed distribution, applying a log transformation might help


library(sandwich)
library(lmtest)
coeftest(fit, vcov = vcovHC(fit, type = "HC1")) # Use heteroscedasticity-robust standard errors, which correct for heteroscedasticity without needing to transform the data.
```

# D. Check for Autocorrelation (Independence of Errors)

Residuals should not be correlated over time or order.

```{r}
library(car)
durbinWatsonTest(fit)
# p > 0.05 → No significant autocorrelation (good).
# p < 0.05 → Autocorrelation detected (consider time-series models).
```

# E. Check for Multicollinearity

High correlation between predictors can distort coefficient estimates

```{r}
library(car)
vif(fit)
# GVIF (Generalized Variance Inflation Factor): Used for categorical variables (factors with multiple levels).
# GVIF^(1/(2*Df)): Adjusted VIF for easier interpretation when a factor has more than one degree of freedom (Df).
# VIF < 5: No severe multicollinearity (good).
# VIF > 10: Strong multicollinearity (consider removing or combining variables).
```

