---
title: "Solution to Neural Network, Classification Task, Exercise"
author: "Your name"
date: "mm/dd/yyyy"
output:
  pdf_document: default
  word_document: default
---

# Load data and packages
```{r}
if (!requireNamespace("tidyverse")) install.packages("tidyverse")
if (!requireNamespace("caret")) install.packages("caret")
if (!requireNamespace("neuralnet")) install.packages("neuralnet")
if (!requireNamespace("e1071")) install.packages("e1071")

library(tidyverse)
library(caret)
library(neuralnet)
library(e1071)
```

```{r}
banknote <- read.csv('banknote.csv')
banknote <- na.omit(banknote)
cat('There are', nrow(banknote), 'observations left.')
```

```{r}
str(banknote)
```

# Question 1

The first question use SSE as the loss function, so we do not factorize the response variable now.

```{r}
set.seed(123)
training.samples <- banknote$class %>% createDataPartition(p = 0.75, list = FALSE)
train.data  <- banknote[training.samples, ]
test.data <- banknote[-training.samples, ]
```

# Question 2

```{r}
set.seed(123)
model <- neuralnet(class~., data = train.data, hidden = 0, err.fct = "sse", linear.output = F)
# If hidden = 0, neuralnet will create a model with only the input and output layers
# Loss function: SSE
# err.fct = function(y, yhat) mean((y - yhat)^2) for MSE
# linear.output = FALSE applies a logistic (sigmoid) activation function to the output.
plot(model, rep = "best")
```

```{r}
probabilities <- model %>% predict(test.data) %>% as.vector()
predicted.classes <- ifelse(probabilities > 0.5, 1, 0)
predicted.classes <- factor(predicted.classes, levels = c(0, 1))
true.classes <- factor(test.data$class, levels = c(0, 1))

cm <- confusionMatrix(predicted.classes, true.classes, positive = "1")
cm
```

# Question 3

```{r}
set.seed(123)
model <- neuralnet(class~., data = train.data, hidden = 0, err.fct = "ce", linear.output = F)
plot(model, rep = "best")
```

```{r}
probabilities <- model %>% predict(test.data)
predicted.classes <- ifelse(probabilities > 0.5, 1, 0)
predicted.classes <- factor(predicted.classes, levels = c(0, 1))
true.classes <- factor(test.data$class, levels = c(0, 1))

cm <- confusionMatrix(predicted.classes, true.classes, positive = "1")
cm
```

# Question 4

```{r}
set.seed(123)
model <- glm(class~., family = binomial, data = train.data)
model
```

```{r}
probabilities <- model %>% predict(test.data, type = 'response')
predicted.classes <- ifelse(probabilities > 0.5, 1, 0)
predicted.classes <- factor(predicted.classes, levels = c(0, 1))
true.classes <- factor(test.data$class, levels = c(0, 1))

cm <- confusionMatrix(predicted.classes, true.classes, positive = "1")
cm
```

# Question 5
```{r}
set.seed(123)
model <- neuralnet(class~., data = train.data, hidden = 3, err.fct = "sse", linear.output = F)
plot(model, rep = "best")

# In neuralnet, you can adjust the initial weights using the startweights argument. 
# This allows you to specify the starting values for the network’s weights instead of relying on the default random initialization.
# E.g we have 4 input features (i.e., p = 4), 3 hidden neurons, and 1 output neuron.
# Total weights=(p+1)×h+(h+1)×1 = (4+1)×3+(3+1)×1=5×3+4×1=15+4 = 19
init_weights <- runif(19, min = -1, max = 1)  # Randomly initialize between -1 and 1
model_w <- neuralnet(class~., data = train.data, hidden = 3, err.fct = "sse",
                   linear.output = FALSE, startweights = init_weights)
plot(model_w, rep = "best")
```

The three neurons in the hidden layer of your neural network share the same activation function, but they learn different feature representations.

All hidden neurons use the same activation function (e.g., sigmoid or tanh).
Each neuron has different weights and biases, so they learn different feature transformations.
The output neuron aggregates their outputs to make the final prediction.


```{r}
probabilities <- model %>% predict(test.data)
predicted.classes <- ifelse(probabilities > 0.5, 1, 0)
predicted.classes <- factor(predicted.classes, levels = c(0, 1))
true.classes <- factor(test.data$class, levels = c(0, 1))

cm <- confusionMatrix(predicted.classes, true.classes, positive = "1")
cm
```

The prediction with hidden layer is better than no hidden layer.

# Question 6

```{r}
set.seed(123)
model <- neuralnet(class~., data = train.data, hidden = 3, err.fct = "ce", linear.output = F)
plot(model, rep = "best")
```
```{r}
probabilities <- model %>% predict(test.data)
predicted.classes <- ifelse(probabilities > 0.5, 1, 0)
predicted.classes <- factor(predicted.classes, levels = c(0, 1))
true.classes <- factor(test.data$class, levels = c(0, 1))

cm <- confusionMatrix(predicted.classes, true.classes, positive = "1")
cm
```

The prediction with hidden layer is better than no hidden layer in this task.
