---
title: "NN_classification_excercies"
output: pdf_document
---

**Using only the training data, try several neural network settings and choose the best neural network classifier for predicting class.**

You may vary:
1. the number of hidden neurons,
2. the loss function ("sse" or "ce"),
3. and the classification cutoff if needed.

Use the training set to select the model, and then evaluate the final chosen model on the testing set. Report:
1. the tuning settings you tried,
2. which model was selected as the best,
3. the confusion matrix on the testing set,
4. the sensitivity, specificity, and overall accuracy.

Briefly explain why your chosen model is the best.

```{r}
if (!requireNamespace("tidyverse")) install.packages("tidyverse")
if (!requireNamespace("caret")) install.packages("caret")
if (!requireNamespace("neuralnet")) install.packages("neuralnet")
if (!requireNamespace("e1071")) install.packages("e1071")
library(tidyverse)
library(caret)
library(neuralnet)
library(e1071)

banknote <- read.csv('banknote.csv')
banknote <- na.omit(banknote)
set.seed(123)
training.samples <- banknote$class %>% createDataPartition(p = 0.75, list = FALSE)
train.data  <- banknote[training.samples, ]
test.data <- banknote[-training.samples, ]
```


```{r}
# Tune the neural network model
set.seed(123)

# -----------------------------------
# Step 1: Standardize predictors
# (IMPORTANT: use training-set statistics only)
# -----------------------------------
preproc <- preProcess(train.data[, -ncol(train.data)], method = c("center", "scale"))

train.x <- predict(preproc, train.data[, -ncol(train.data)])
test.x  <- predict(preproc, test.data[, -ncol(test.data)])

train.data.scaled <- data.frame(train.x, class = train.data$class)
test.data.scaled  <- data.frame(test.x, class = test.data$class)

# -----------------------------------
# Step 2: Split training data into sub-training and validation sets
# -----------------------------------
sub_train_index <- createDataPartition(train.data.scaled$class, p = 0.8, list = FALSE)
sub_train <- train.data.scaled[sub_train_index, ]
valid_data <- train.data.scaled[-sub_train_index, ]

# -----------------------------------
# Step 3: Function to fit model and return validation accuracy
# Use tryCatch to avoid crashing if one model fails
# -----------------------------------
get_valid_accuracy <- function(hidden_units, loss_type) {
  set.seed(123)
  
  out <- tryCatch({
    fit <- neuralnet(
      class ~ .,
      data = sub_train,
      hidden = hidden_units,
      err.fct = loss_type,
      linear.output = FALSE,
      stepmax = 1e6
    )
    
    prob <- predict(fit, valid_data) |> as.vector()
    pred <- ifelse(prob > 0.5, 1, 0)
    
    pred <- factor(pred, levels = c(0, 1))
    truth <- factor(valid_data$class, levels = c(0, 1))
    
    cm <- confusionMatrix(pred, truth, positive = "1")
    
    data.frame(
      hidden = hidden_units,
      loss = loss_type,
      valid_accuracy = as.numeric(cm$overall["Accuracy"]),
      status = "success"
    )
  }, error = function(e) {
    data.frame(
      hidden = hidden_units,
      loss = loss_type,
      valid_accuracy = NA,
      status = "failed"
    )
  })
  
  return(out)
}

# -----------------------------------
# Step 4: Grid of tuning parameters
# -----------------------------------
tuning_grid <- expand.grid(
  hidden = c(0, 1, 3, 5),
  loss = c("sse", "ce")
)

# -----------------------------------
# Step 5: Try all settings
# -----------------------------------
results <- purrr::map2_df(tuning_grid$hidden, tuning_grid$loss, get_valid_accuracy)
results

# Keep only successful fits
results_clean <- results %>%
  filter(status == "success") %>%
  arrange(desc(valid_accuracy))

results_clean

# -----------------------------------
# Step 6: Select the best setting
# -----------------------------------
best_setting <- results_clean %>%
  slice(1)

best_setting

# -----------------------------------
# Step 7: Refit the best model using the FULL scaled training data
# -----------------------------------
set.seed(123)
best_model <- neuralnet(
  class ~ .,
  data = train.data.scaled,
  hidden = best_setting$hidden,
  err.fct = as.character(best_setting$loss),
  linear.output = FALSE,
  stepmax = 1e6
)

plot(best_model, rep = "best")

# -----------------------------------
# Step 8: Evaluate on the scaled testing data
# -----------------------------------
best_prob <- predict(best_model, test.data.scaled) |> as.vector()
best_pred <- ifelse(best_prob > 0.5, 1, 0)

best_pred <- factor(best_pred, levels = c(0, 1))
true_class <- factor(test.data.scaled$class, levels = c(0, 1))

best_cm <- confusionMatrix(best_pred, true_class, positive = "1")
best_cm

# -----------------------------------
# Step 9: Report metrics
# -----------------------------------
data.frame(
  Sensitivity = best_cm$byClass["Sensitivity"],
  Specificity = best_cm$byClass["Specificity"],
  Accuracy = best_cm$overall["Accuracy"]
)

```