---
title: "Solution to the Logistic Regression Example"
author: "Your name"
date: "02/16/2025"
output:
  pdf_document: default
  word_document: default
---

# Install packages and load data

R does not know which CRAN repository to use when installing packages during knitting.
```{r}
options(repos = c(CRAN = "https://cloud.r-project.org/"))
install.packages('mlbench')

if (!requireNamespace("mlbench")) install.packages('mlbench')
library(mlbench)
library(tidyverse)
library(caret)
```
Data Description

The 9 variables of the pimaindiansdiabetes2 tbl data frame are the following ones:  
  
pregnant Number of times pregnant (numeric);  
glucose Plasma glucose concentration a 2 hours in an oral glucose tolerance test (numeric);  
pressure Diastolic blood pressure (mm Hg) (numeric);  
triceps Triceps skin fold thickness (mm) (numeric);  
insulin 2-Hour serum insulin (mu U/ml)(numeric);  
mass Body mass index (weight in kg/(height in m)^2) (numeric);  
pedigree Diabetes pedigree function (numeric);  
age Age (years) (numeric);  
diabetes class variable: diabetic or not (factor with 2 levels: neg and pos)  

```{r}
data("PimaIndiansDiabetes2", package = "mlbench")
PimaIndiansDiabetes2 <- na.omit(PimaIndiansDiabetes2) # remove missing values
# Inspect the data
sample_n(PimaIndiansDiabetes2, 3)
# Split the data into training and test set
```


# Question (a)
```{r}
set.seed(123)
training.samples <- PimaIndiansDiabetes2$diabetes %>% 
  createDataPartition(p = 0.8, list = FALSE)
train.data  <- PimaIndiansDiabetes2[training.samples, ]
test.data <- PimaIndiansDiabetes2[-training.samples, ]
```


# Question (b)
```{r}
# Fit the model
model <- glm( diabetes ~., data = train.data, family = binomial)
# Summarize the model
summary(model)
```

# Question (c)

## prediction
```{r}
probabilities <- model %>% predict(test.data, type = "response")
predicted.classes <- ifelse(probabilities > 0.5, "pos", "neg")
```

## confusion matrix
```{r}
table(predicted.classes,test.data$diabetes)
```


## c(i)
```{r}
# Model accuracy
mean(predicted.classes == test.data$diabetes)
```

```{r}
(44+15)/(44+11+8+15)
```

overall accuracy = P(predicted correctly) = (44+15)/(44+11+8+15) = 0.7564


## c(ii)

```{r}
sum((predicted.classes == 'pos') * (test.data$diabetes == 'pos')) / sum(test.data$diabetes == 'pos')
```
```{r}
15/(11+15)
```

Sensitivity = P(predicted positive | positive) = 15/(11+15) = 0.5769

## c(iii)

```{r}
sum((predicted.classes == 'neg') * (test.data$diabetes == 'neg')) / sum(test.data$diabetes == 'neg')
```
```{r}
44/(44+8)
```

Specificity = P(predicted negative | negative) = 44/(44+8) = 0.8462