---
title: "Solution to multiple regression example"
author: "Your name"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

# Install packages 

```{r}
library(MASS)
library(tidyverse)
library(caret)
library(glmnet)
```

# Question 1
```{r}
data1 <- read.csv('math.csv', sep = ';')
data1 <- subset(data1, select = - c(G1,G2))
set.seed(123)
training.samples <- data1$G3 %>% createDataPartition(p = 0.75, list = FALSE) # caret pkg
train.data  <- data1[training.samples, ]
test.data <- data1[-training.samples, ]
```

# Question 2
```{r}
fit <- lm(G3~., data = train.data)
fit_step <- stepAIC(fit, k = log(nrow(train.data)), trace = 0) # MASS pkg
pred <- fit_step %>% predict(test.data)
data.frame(
RMSE = RMSE(pred, test.data$G3),
Rsquare = R2(pred, test.data$G3)
)
```

# Question 3
```{r}
library(leaps)
fit_bs <- regsubsets(G3~., data = train.data, nvmax = 30) # leaps pkg
result <- summary(fit_bs)
which.min(result$rss)
```
```{r}
result$which[30,]
```

```{r}
fit_bs <- lm(G3~., data = train.data)
pred <- fit_bs %>% predict(test.data)
data.frame(
RMSE = RMSE(pred, test.data$G3),
Rsquare = R2(pred, test.data$G3)
)
```

# Question 4

## Two models with different predictors, three predictors each
## BICs: Smaller -> better
```{r}
mod_step <- lm(G3~Medu +  failures + romantic, data = train.data)
mod_bs <- lm(G3~., data = train.data)
BIC(mod_step) # better
BIC(mod_bs) 

pred_step = predict(mod_step, test.data)
pred_bs = predict(mod_bs, test.data)
rmse_step = sqrt(mean((pred_step - test.data$G3)^2))
rmse_bs = sqrt(mean((pred_bs - test.data$G3)^2))
print(paste("step RMSE:", rmse_step))
print(paste("best subset RMSE:", rmse_bs))

r_squared_step = cor(test.data$G3, pred_step)^2
r_squared_bs = cor(test.data$G3, pred_bs)^2
print(paste("step R^2:", r_squared_step))
print(paste("best subset R^2:", r_squared_bs))
```
Based on the BIC values of the fitted model using the training data - the one selected by the Stepwise regression using the BIC criterion has a smaller BIC value. This is because the Stepwise procedure uses the BIC critrion while the Best Subset procedure here uses the SSE (RSS) as selection criterion. 

According to the RMSE and R2 of the predition for the test data, the Best Subset one is better since R2 always increases (or stays the same) when adding more predictors, this method selected all 30 predictors, even if some of them contribute little predictive power or cause overfitting. The lower RMSE on the testing set suggests that the additional predictors helped capture more variance, but this might be due to chance (overfitting risk).

modifications: 
1. Instead of using a single test set RMSE, perform k-fold cross-validation to assess generalizability. This helps to determine whether the extra predictors genuinely improve performance or are just noise.
2. use adjusted R2
3. try to do model diagnosis do see if transformation is needed.