---
title: "Stepwise and best subset variable selection"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r}
library(MASS)
library(tidyverse)
library(caret)
library(glmnet)
library(leaps)

# Use variables except G1 and G2 to predicto G3
# y: G3
# x: all the other variables
data <- read.csv('math.csv', sep = ';')
data <- subset(data, select = - c(G1,G2))
str(data)

set.seed(123)
training.samples <- data$G3 %>% createDataPartition(p = 0.75, list = FALSE) # caret pkg
# Uses createDataPartition() from the caret package to split the data into training (75%) and test (25%) sets.
# The %>% operator (pipe) is from the tidyverse pkg
train.data  <- data[training.samples, ]
test.data <- data[-training.samples, ]
```

## stepwise model selection using the Bayesian Information Criterion (BIC)



AIC=−2logL+2k

L = Maximum likelihood of the model
k = Number of parameters in the model

BIC=−2logL+klogn

n = Number of observations

Backward:
1. Starts with the full model (all predictors).
2. Removes the least significant predictor (one with the highest p-value) iteratively.
3. Stops when removing predictors no longer improves BIC.
4. Returns the final optimal model (fit_step).

```{r}
fit <- lm(G3~., data = train.data) # a full model
fit_step <- stepAIC(fit, k = log(nrow(train.data)), trace = 1) # MASS pkg
# stepwise selection to find the best subset of predictors step by step.
# backward selection starts with all predictors and iteratively removes the least important one.
# k = 2: AIC
# k = log(n) BIC
# trace = 1: print the selection process. Larger values may give more information on the fitting process.
# direction = 'both'/'backward'/'forward'
pred <- fit_step %>% predict(test.data)
data.frame(
RMSE = RMSE(pred, test.data$G3),
Rsquare = R2(pred, test.data$G3)
)
```

forward:
1. Starts with the null model (intercept only).
2. Adds the most significant predictor iteratively.
3. Stops when removing predictors no longer improves BIC.
4. Returns the final optimal model (fit_step).

```{r}
null_model <- lm(G3 ~ 1, data = train.data)
full_model <- lm(G3 ~ ., data = train.data)
fit_step_fwd <- stepAIC(null_model, scope = list(lower = null_model, upper = full_model), 
                        direction = "forward", k = log(nrow(train.data)), trace = 1)
pred_fwd <- fit_step_fwd %>% predict(test.data)
data.frame(
RMSE = RMSE(pred_fwd, test.data$G3),
Rsquare = R2(pred_fwd, test.data$G3)
)
```

Bidirectional (Both)

```{r}
fit_step_both <- stepAIC(fit, k = log(nrow(train.data)), direction = "both", trace = 1)
pred_both <- fit_step_both %>% predict(test.data)
data.frame(
RMSE = RMSE(pred_both, test.data$G3),
Rsquare = R2(pred_both, test.data$G3)
)
```

Backward	
Full model (G3 ~ .)	Removes least significant predictors iteratively	
When you have many predictors and want to simplify the model

Forward	
Intercept-only (G3 ~ 1)	Adds most significant predictors iteratively	
When you have few predictors and want to build a model from scratch

Both Directions	
Either (G3 ~ . or G3 ~ 1)	Adds and removes predictors iteratively	
When unsure whether to start with a full or empty model


## These three stepwise regression methods (forward, backward, and bidirectional) may not always give the same final model. 
1. Stepwise methods are greedy algorithms, meaning they make the best choice at each step but do not guarantee the best overall model.
2. The sequence of adding/removing predictors can affect which predictors remain in the final model.
3. Backward selection removes predictors based on p-values, but if two variables are correlated, removing one may make the other significant.

## Bestsubset Variable Selection

It selects the best set of predictors based on BIC (or AIC or R square)
Tries all possible combinations of predictors to find the best model
# Global Optima!
Unlike stepwise selection, which follows a specific path (adding/removing one variable at a time), best subset selection evaluates all possible models.

```{r}
fit_bs <- regsubsets(G3 ~ ., data = train.data, nvmax = 30)
# nvmax = 30: sets the maximum number of predictors to consider.
result <- summary(fit_bs)
which.min(result$bic)
result$which[3,]

n <- nrow(train.data)  # Sample size
p <- 1:(length(result$rss))  # Model sizes (number of predictors + intercept)
aic_values <- n * log(result$rss / n) + 2 * p  # Compute AIC for each model
which.min(aic_values)  # Get the model size with the lowest AIC
result$which[12,]


which.max(result$adjr2)
result$which[1,]
```


# Best Subset Selection (regsubsets())	
Tries all possible combinations of predictors	
Finds the truly best model	
Computationally expensive for many predictors

# Stepwise Selection (stepAIC())	
Adds/removes one predictor at a time	
Faster, works well for moderate predictor sets	
Can miss the best model due to greedy approach
