---
title: "Transformer Encoder - Standard Solutions (Revised)"
output:
  pdf_document:
    toc: true
    toc_depth: '2'
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, message=FALSE, warning=FALSE}
library(keras3)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)
set.seed(123)

# This Rmd uses keras3 and therefore requires a Python backend.
# Example setup (run once if needed):
# keras3::install_keras(backend = "tensorflow")

make_sequences <- function(df, feature_cols, target_col, seq_len) {
  X_raw <- as.matrix(df[, feature_cols])
  y_raw <- df[[target_col]]
  n <- nrow(df)
  n_samples <- n - seq_len
  n_features <- length(feature_cols)
  X <- array(NA_real_, dim = c(n_samples, seq_len, n_features))
  y <- numeric(n_samples)
  for (i in 1:n_samples) {
    X[i, , ] <- X_raw[i:(i + seq_len - 1), ]
    y[i] <- y_raw[i + seq_len]
  }
  list(X = X, y = y)
}

# keras3 + reticulate can sometimes reject arrays with an unexpected backend type.
# This helper forces clean float32 tensors before model fitting.
make_keras_input <- function(X, y = NULL) {
  X_dim <- as.integer(dim(X))
  X_clean <- array(as.numeric(X), dim = X_dim)
  storage.mode(X_clean) <- "double"
  
  if (is.null(y)) {
    return(X_clean)
  }
  
  y_clean <- array(as.numeric(y), dim = c(as.integer(length(y)), 1L))
  storage.mode(y_clean) <- "double"
  
  list(X = X_clean, y = y_clean)
}

inverse_signal <- function(x_scaled, mean_signal, sd_signal) {
  as.numeric(x_scaled) * sd_signal + mean_signal
}

recursive_forecast <- function(model, last_window, h = 5) {
  # last_window is a matrix with shape [seq_len, n_features]
  current_window <- last_window
  preds <- numeric(h)
  for (i in 1:h) {
    x_input <- op_convert_to_tensor(
      array(as.numeric(current_window), dim = c(1, nrow(current_window), ncol(current_window))),
      dtype = "float32"
    )
    next_pred <- as.numeric(predict(model, x_input, verbose = 0))
    preds[i] <- next_pred

    # For this exercise, future exogenous variables are held at their last observed values.
    new_row <- current_window[nrow(current_window), ]
    new_row[1] <- next_pred  # first feature is signal
    current_window <- rbind(current_window[-1, ], new_row)
  }
  preds
}
```

# 1. Read the dataset and plot signal versus time

```{r}
data <- read_csv("synthetic_transformer_series.csv", show_col_types = FALSE)

ggplot(data, aes(x = time, y = signal)) +
  geom_line(linewidth = 0.7, color = "steelblue") +
  labs(title = "Signal over Time", x = "Time", y = "Signal") +
  theme_minimal()
```

```{r}
# Chronological split: first 75% train, next 15% validation, last 10% test.
n <- nrow(data)
train_end <- floor(0.75 * n)
val_end <- floor(0.90 * n)

train_df <- data[1:train_end, ]
val_df <- data[(train_end + 1):val_end, ]
test_df <- data[(val_end + 1):n, ]

nrow(train_df)
nrow(val_df)
nrow(test_df)
```

# 2. Create 3D predictor arrays and responses

```{r}
seq_len <- 20
feature_cols <- c("signal", "exog1", "exog2")
target_col <- "signal"

# Standardize features using training-set statistics only.
train_means <- sapply(train_df[, feature_cols], mean)
train_sds <- sapply(train_df[, feature_cols], sd)
mean_signal <- train_means[["signal"]]
sd_signal <- train_sds[["signal"]]

scale_with_train <- function(df, means, sds, cols) {
  out <- df
  out[, cols] <- scale(df[, cols], center = means, scale = sds)
  out
}

train_scaled <- scale_with_train(train_df, train_means, train_sds, feature_cols)
val_scaled <- scale_with_train(val_df, train_means, train_sds, feature_cols)
test_scaled <- scale_with_train(test_df, train_means, train_sds, feature_cols)

# To create validation and testing sequences, prepend the needed history.
val_input_df <- bind_rows(tail(train_scaled, seq_len), val_scaled)
test_input_df <- bind_rows(tail(bind_rows(train_scaled, val_scaled), seq_len), test_scaled)

train_seq <- make_sequences(train_scaled, feature_cols, target_col, seq_len)
val_seq <- make_sequences(val_input_df, feature_cols, target_col, seq_len)
test_seq <- make_sequences(test_input_df, feature_cols, target_col, seq_len)

dim(train_seq$X)
dim(val_seq$X)
dim(test_seq$X)
```

```{r}
# Convert arrays/vectors to clean float32 tensors before fitting.
train_keras <- make_keras_input(train_seq$X, train_seq$y)
val_keras <- make_keras_input(val_seq$X, val_seq$y)
test_keras_X <- make_keras_input(test_seq$X)
```

# 3. LSTM baseline

```{r}
# One LSTM layer with 32 hidden units.
# We use an Input layer first because this is the recommended keras3 style.
inputs_lstm <- layer_input(shape = c(seq_len, length(feature_cols)))
outputs_lstm <- inputs_lstm |>
  layer_lstm(units = 32) |>
  layer_dense(units = 1)

model_lstm <- keras_model(inputs = inputs_lstm, outputs = outputs_lstm)
model_lstm$compile(
  optimizer = optimizer_adam(),
  loss = "mse",
  metrics = list("mse")
)

callback_es <- callback_early_stopping(
  monitor = "val_loss",
  patience = 10,
  restore_best_weights = TRUE
)

history_lstm <- model_lstm$fit(
  x = train_keras$X,
  y = train_keras$y,
  validation_data = list(val_keras$X, val_keras$y),
  epochs = as.integer(80),
  batch_size = as.integer(32),
  callbacks = list(callback_es),
  verbose = 0
)
```

```{r}
extract_history_df <- function(history_obj) {
  # Try several possible storage locations used by keras / keras3 / reticulate
  candidates <- list(
    history_obj$history,
    history_obj$params,
    history_obj
  )
  
  hist_list <- NULL
  
  for (obj in candidates) {
    if (is.null(obj)) next
    
    # case 1: already a list and contains loss
    if (is.list(obj) && !is.environment(obj) && "loss" %in% names(obj)) {
      hist_list <- obj
      break
    }
    
    # case 2: environment containing loss / val_loss
    if (is.environment(obj)) {
      env_names <- ls(obj)
      if ("loss" %in% env_names) {
        hist_list <- mget(env_names, envir = obj)
        break
      }
    }
  }
  
  if (is.null(hist_list)) {
    stop("Could not extract loss history from the History object. Please run str(history_lstm).")
  }
  
  if (!"loss" %in% names(hist_list)) {
    stop("Training loss not found in history object.")
  }
  
  if (!"val_loss" %in% names(hist_list)) {
    warning("Validation loss not found; plotting training loss only.")
    out <- data.frame(
      epoch = seq_along(unlist(hist_list$loss)),
      training = as.numeric(unlist(hist_list$loss))
    )
    return(out)
  }
  
  data.frame(
    epoch = seq_along(unlist(hist_list$loss)),
    training = as.numeric(unlist(hist_list$loss)),
    validation = as.numeric(unlist(hist_list$val_loss))
  )
}


loss_lstm_wide <- extract_history_df(history_lstm)

loss_lstm <- tidyr::pivot_longer(
  loss_lstm_wide,
  cols = c(training, validation),
  names_to = "series",
  values_to = "loss"
)

ggplot(loss_lstm, aes(x = epoch, y = loss, color = series)) +
  geom_line(linewidth = 1) +
  labs(
    title = "LSTM Training and Validation Loss",
    x = "Epoch",
    y = "MSE Loss",
    color = NULL
  ) +
  theme_minimal()
```

# 4. Transformer encoder model

```{r}
# A compact Transformer encoder for time-series regression.
# num_heads = 2 controls how many attention heads are used.
# key_dim = 16 sets the latent dimension inside each attention head.
# ff_dim = 32 controls the width of the feed-forward subnetwork.
# dropout = 0.1 adds regularization.
inputs_tr <- layer_input(shape = c(seq_len, length(feature_cols)))

attn_out <- layer_multi_head_attention(
  num_heads = 2,
  key_dim = 16,
  dropout = 0.1
)(inputs_tr, inputs_tr)

x <- layer_add(list(inputs_tr, attn_out)) |>
  layer_layer_normalization(epsilon = 1e-6)

ffn <- x |>
  layer_dense(units = 32, activation = "relu") |>
  layer_dense(units = length(feature_cols))

x <- layer_add(list(x, ffn)) |>
  layer_layer_normalization(epsilon = 1e-6) |>
  layer_global_average_pooling_1d() |>
  layer_dropout(rate = 0.1)

outputs_tr <- x |>
  layer_dense(units = 16, activation = "relu") |>
  layer_dense(units = 1)

model_tr <- keras_model(inputs = inputs_tr, outputs = outputs_tr)
model_tr$compile(
  optimizer = optimizer_adam(),
  loss = "mse",
  metrics = list("mse")
)

history_tr <- model_tr$fit(
  x = train_keras$X,
  y = train_keras$y,
  validation_data = list(val_keras$X, val_keras$y),
  epochs = as.integer(80),
  batch_size = as.integer(32),
  callbacks = list(callback_es),
  verbose = 0
)
```

```{r}
loss_tr_wide <- extract_history_df(history_tr)

loss_tr <- tidyr::pivot_longer(
  loss_tr_wide,
  cols = c(training, validation),
  names_to = "series",
  values_to = "loss"
)

ggplot(loss_tr, aes(x = epoch, y = loss, color = series)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Transformer Training and Validation Loss",
    x = "Epoch",
    y = "MSE Loss",
    color = NULL
  ) +
  theme_minimal()
```

# 5. Test MSE and prediction plot

```{r}
pred_lstm_scaled <- as.numeric(predict(model_lstm, test_keras_X, verbose = 0))
pred_tr_scaled <- as.numeric(predict(model_tr, test_keras_X, verbose = 0))
actual_test_scaled <- test_seq$y

mse_lstm <- mean((pred_lstm_scaled - actual_test_scaled)^2)
mse_tr <- mean((pred_tr_scaled - actual_test_scaled)^2)

mse_lstm
mse_tr
```

```{r}
actual_test <- inverse_signal(actual_test_scaled, mean_signal, sd_signal)
pred_lstm <- inverse_signal(pred_lstm_scaled, mean_signal, sd_signal)
pred_tr <- inverse_signal(pred_tr_scaled, mean_signal, sd_signal)

plot_df <- data.frame(
  step = 1:length(actual_test),
  Actual = actual_test,
  LSTM = pred_lstm,
  Transformer = pred_tr
) |>
  pivot_longer(cols = c(Actual, LSTM, Transformer), names_to = "series", values_to = "value")

ggplot(plot_df, aes(x = step, y = value, color = series)) +
  geom_line(linewidth = 0.8) +
  labs(title = "Testing Predictions: LSTM vs Transformer", x = "Testing Step", y = "Signal") +
  theme_minimal()
```

# 6. Which model performs better?

LSTM

In this example, the Transformer appears to underfit. Its predictions collapse toward a smooth mean-like trajectory and fail to track local turning points. A likely reason is that the dataset is relatively small, and the current encoder design with global average pooling may wash out the most recent time-step information that is important for one-step-ahead forecasting.

# 7. Recursive 5-step forecast using the better model

```{r}
better_model <- if (mse_tr < mse_lstm) model_tr else model_lstm
model_name <- if (mse_tr < mse_lstm) "Transformer" else "LSTM"

last_window <- as.matrix(tail(test_scaled[, feature_cols], seq_len))
future_scaled <- recursive_forecast(better_model, last_window, h = 5)
future_signal <- inverse_signal(future_scaled, mean_signal, sd_signal)
future_steps <- (max(data$time) + 1):(max(data$time) + 5)

history_plot <- tail(data[, c("time", "signal")], 30)
history_plot$series <- "Observed"

future_plot <- data.frame(
  time = future_steps,
  signal = future_signal,
  series = "Forecast"
)

plot_future <- bind_rows(history_plot, future_plot)

ggplot(plot_future, aes(x = time, y = signal, color = series)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  labs(
    title = paste("Recursive 5-Step Forecast Using", model_name),
    x = "Time",
    y = "Signal"
  ) +
  theme_minimal()
```
